# Appendix {-}

Table \@ref(tab:TD-tab) shows exact power for the Test of Deficit [TD: @crawford_comparing_1998], for a range of likely parameters. Power for the TD is calculated from the formula:

\begin{equation*}
power = 1 - \beta = T_{n-1}\left(t_{\alpha, \ n-1} \Bigg\rvert \frac{x^* - \overline{x}}{\sigma \sqrt{\frac{n+1}{n}}}\right)
\end{equation*}

Where $T_{n-1}(. \rvert \theta)$ is the cumulative distribution function for the
non-central t-distribution with $n-1$ degrees of freedom and non-centrality
parameter $TD = \frac{x^* - \overline{x}}{\sigma \sqrt{\frac{n+1}{n}}}$
is the $\alpha$ quantile of the *central* t-distribution on $n-1$ degrees of
freedom. See @harrison_sample_2004 for discussion on using the non-central
t-distribution for power calculations for t-tests.

Note that the Bayesian Test of Deficit [@crawford_comparing_2011] produces
equivalent results to the TD, and has frequentist properties, so may be
considered to have the same effective levels of power as the TD. Note also that
the Unstandardised Difference Test (UDT) [@crawford_testing_2005] is
identical to applying the TD to between-task discrepancy scores. Power for the
UDT is thus identical with power the TD, but related to the effect size of a
between-task discrepancy ($Z_{DCC}$) rather than the effect size of a deficit ($Z_{CC}$).

Table \@ref(tab:RSDT-tab) shows estimated power for the Revised Standardised Difference Test
[RSDT: @crawford_testing_2005], for a range of likely parameters. Power for
the RSDT was estimated using Monte Carlo simulation, by generating n+1 random
draws from a bivariate normal distribution, imposing the desired ZDCC on the
n+1th draw and comparing that observation to the n others using the RSDT test
formula. This was repeated 1 000 000 times, and power was found by dividing the
number of significant tests by the total number of tests run, for each parameter
combination of interest. The code for our calculations and simulations is
available 
[here](https://github.com/jorittmo/neuropsychology_power_primer)^[https://github.com/jorittmo/neuropsychology_power_primer].
Power calculations for a wider range of single-case tests
are supported by the singcar package for R, the developmental version of which can be 
accessed [here](https://github.com/jorittmo/singcar)^[https://github.com/jorittmo/singcar].


\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}

\setlength{\parindent}{0in}
\setlength{\leftskip}{-0.5in}
\setlength{\parskip}{0pt}



\newgeometry{margin=1in}

(ref:TD-tab-text) *Note:* $\text{Z}_{\text{CC}} = \frac{x^* - \overline{x}}{s_x}$, i.e. the standardised case score for the task of interest.

(ref:TD-tab-cap) Power for the Test of Deficit [@crawford_comparing_1998].


```{r TD-tab, echo=FALSE, message=FALSE}

library(kableExtra)

ana_TD_pwr <- function () {
  xbar <- 0
  sigma <- 1
  ana_TD_pwr_data <- data.frame(deficit = double(), ncon = double(), power = double())
  for (n in seq(4, 24, by = 2)) {
    for (alpha in c(0.05, 0.1)) {
      for (tail in c("Two", "One")) {
        
        if (tail == "One") {
          pd <- data.frame(deficit = seq(-4, -0.5, by = 0.5))%>%
            mutate(Tail = tail) %>% 
            mutate(alpha = alpha) %>% 
            mutate(ncon = n) %>%
            mutate(power = pt(qt(alpha, df = n-1,
                                 lower.tail = T),
                              ncp = ((deficit-xbar)/(sigma*sqrt((n+1)/n))),
                              df = n-1,
                              lower.tail = T
            )) 
                  
                  
                  
        }
        if (tail == "Two") {
          pd <- data.frame(deficit = seq(-4, -0.5, by = 0.5))%>%
            mutate(Tail = tail) %>% 
            mutate(alpha = alpha) %>% 
            mutate(ncon = n) %>%
            mutate(power = pt(qt(alpha/2, df = n-1,
                                 lower.tail = T),
                              ncp = ((deficit-xbar)/(sigma*sqrt((n+1)/n))),
                              df = n-1,
                              lower.tail = T) - pt(-qt(alpha/2, df = n-1,
                                                           lower.tail = T),
                                                       ncp = ((deficit-xbar)/(sigma*sqrt((n+1)/n))),
                                                       df = n-1,
                                                       lower.tail = T) + 1) 
          
        }
        
        ana_TD_pwr_data <- rbind(ana_TD_pwr_data, pd)
      }
    }
  }
  ana_TD_pwr_data
}
ana_TD <- ana_TD_pwr()
  
ana_TD %>% 
  mutate(power = power*100) %>%
  mutate(zcc = deficit*-1) %>%
  select(Tail, alpha, zcc, ncon, power) %>% 
  pivot_wider(names_from = ncon, values_from = power)  %>% 
  arrange(zcc) %>% 
  arrange(alpha) %>% 
  arrange(Tail) -> ana_TD_arr


kable(ana_TD_arr, align = "r", digits = 2, booktabs = T, 
      caption = "(ref:TD-tab-cap)", 
      format = "latex",
      col.names = c("Tails", "(ref:alph)", "Z\\textsubscript{CC}",  unique(ana_TD$ncon)), escape = F) %>%
  kable_styling(latex_options = c("HOLD_position"),  position = "center") %>% 
  collapse_rows(columns= c(1, 2), valign = "top", latex_hline = "custom", custom_latex_hline = c(1, 2)) %>% 
  add_header_above(c(" " = 3, "Control sample size" = 11 )) %>% 
  add_footnote(label = c('(ref:TD-tab-text)'), notation = "none", escape = FALSE)


```


(ref:RSDT-tab-text1) *Note:* $\text{Z}_{\text{DCC}} = \frac{z_x^* -z_y^*}{\sqrt{2-2r}}$, i.e. the standardised case score for the intertask \linebreak discrepancy of
(ref:RSDT-tab-text2) interest where $z_x^*$ and $z_y^*$ are the standardised case scores for the tasks of interest. Higher
(ref:RSDT-tab-text3) power for small samples and small effect sizes is due to loss of control of Type I errors
(ref:RSDT-tab-text4) (Crawford & Garthwaite, 2005). Power calculated with 1 000 000 Monte Carlo simulations.
(ref:alph) $\alpha$

(ref:RSDT-tab-cap) Power for the Revised Standardised Difference Test (Crawford & Garthwaite, 2005). 
 

```{r RSDT-tab, echo=FALSE, message=FALSE}


load("data/RSDT_tab1M.Rdata")

pwr_RSDT_tab %>% 
  mutate(power = power*100) %>% 
  mutate(zdcc = xa_ast*-1) %>%
  mutate(tail = recode(tail, two.sided = "Two", less = "One")) %>% 
  rename(Tail = tail) %>% 
  select(Tail, alpha, zdcc, ncon, power) %>% 
  pivot_wider(names_from = ncon, values_from = power) %>%
  arrange(zdcc) %>% 
  arrange(alpha) %>%
  arrange(Tail) -> rsdt_pwr_wide

kable(rsdt_pwr_wide, align = "r", digits = 2, booktabs = T, 
      caption = "(ref:RSDT-tab-cap)", 
      format = "latex",
      col.names = c( "Tails", "(ref:alph)", "Z\\textsubscript{DCC}", unique(pwr_RSDT_tab$ncon)), escape = F) %>%
  kable_styling(latex_options = c("HOLD_position", "scale_down")) %>% 
  collapse_rows(columns= c(1, 2), valign = "top", latex_hline = "custom", custom_latex_hline = c(1, 2)) %>% 
  add_header_above(c(" " = 3, "Control sample size" = 11 )) %>%
  add_footnote(label = c('(ref:RSDT-tab-text1)', '(ref:RSDT-tab-text2)', '(ref:RSDT-tab-text3)',
                         '(ref:RSDT-tab-text4)'),
               notation = "none", escape = FALSE)


```



\clearpage
\restoregeometry







